\chapter{Preface}
\lettrine{T}{his} text accompanies the course ``Mathematical Analysis 2'' taught at the University of Rome Tor Vergata in the department of engineering for the academic year 2021/22.
The course was led by Oliver Butterley, in collaboration with Alessio Ranallo. 

The aim of this document is to concisely describe the fundamental details related to the material of the course.
They are aptly named as ``notes'' and are most likely not the comprehensive source of all relevant information.
We have easy access to a huge volume of resources and so here we will make connections to whatever is useful, whenever we can. 

These notes are merely written text whereas the central part of the course remains the time spent working with the material, be it doing exercises, discussing, doing calculations, etc. This is not text for memorising, this is text that aims to help us practice and become stronger thinkers.

This text is freely\footnote{Free both in the sense of \href{https://en.wikipedia.org/wiki/Gratis_versus_libre}{``free speech'' and ``free beer''}.} available at \href{https://github.com/oliver-butterley/ma2}{\textbf{github.com/oliver-butterley/ma2}}.
Everyone is encouraged to contribute improvements to the document during the progress of the course. 

Some of the text comes from previous years and from many other sources, some of the text came to be during the course.
The current version is the product of many people, in particular everyone who has made suggestions in class and pointed out errors or imprecisions and to everyone who suggested useful additional content.


% \cleardoublepage   
% % Make sure contents page starts on right-side page